---
title: "seminrExtras"
description: |
  Advanced modeling and analysis tools that extend SEMinR.
output:
  distill::distill_article:
    toc: true
    toc_depth: 2
---

[**seminrExtras**](https://github.com/sem-in-r/seminrExtras) is a supplementary
package for [SEMinR](index.html) — not a standalone package. It adds advanced
tools for evaluating and validating PLS-SEM models estimated with SEMinR. Every
function accepts an estimated `seminr` model object and returns results with
`print`, `summary`, and `plot` methods.

It is also the home of the example models used in the *Partial Least Squares
Structural Equation Modeling (PLS-SEM) using R* workbook (Hair et al., 2026).

## Installation

**Install the released version from CRAN:**

```{r echo=TRUE, eval=FALSE}
install.packages("seminrExtras")
```

**Or install the development version from GitHub:**

```{r echo=TRUE, eval=FALSE}
# install.packages("remotes")
remotes::install_github("sem-in-r/seminrExtras")
```

Then load it alongside SEMinR:

```{r echo=TRUE, eval=FALSE}
library(seminr)
library(seminrExtras)
```

## What it provides

seminrExtras implements a range of methods for advanced PLS-SEM assessment:

- **Composite Overfit Analysis (COA)** — detect observation-level overfitting via predictive deviance trees and parameter instability analysis (Danks, Ray, & Shmueli, 2024).
- **Congruence testing** — bootstrapped congruence coefficient testing for construct validity (Franke, Sarstedt, & Danks, 2021).
- **Cross-Validated Predictive Ability Test (CVPAT)** — compare a model's predictive performance against benchmarks or alternative models (Liengaard et al., 2021; Sharma et al., 2023).
- **Predictive Contribution of the Mediator (PCM)** — test whether a mediating construct improves out-of-sample predictive accuracy (Danks, 2021).
- **Importance-Performance Map Analysis (IPMA & cIPMA)** — identify constructs that are both important and, via Necessary Condition Analysis integration, necessary (Ringle & Sarstedt, 2016; Hauff et al., 2024).
- **Necessary Condition Analysis (NCA & NCA-ESSE)** — test whether predictors are necessary conditions for an outcome, complementing PLS-SEM's sufficiency logic, with an effect-size sensitivity extension (Dul, 2016; Becker et al., 2026).
- **Confirmatory Tetrad Analysis (CTA-PLS)** — empirically test whether a construct's measurement model is reflective or formative (Gudergan et al., 2008).
- **Unobserved heterogeneity** — detect latent classes via **FIMIX-PLS** (EM-based probabilistic segmentation) and **PLS-POS** (deterministic prediction-oriented segmentation).

## Function reference

| Function | Description |
|---|---|
| `assess_cvpat()` | CVPAT against LM and IA benchmarks |
| `assess_cvpat_compare()` | Compare predictive loss of two PLS models |
| `assess_pcm()` | Predictive Contribution of the Mediator |
| `assess_ipma()` | Importance-Performance Map Analysis (IPMA) |
| `assess_cipma()` | Combined IPMA with Necessary Condition Analysis |
| `assess_coa()` | Composite Overfit Analysis (full pipeline) |
| `predictive_deviance()` | Compute predictive deviance scores |
| `deviance_tree()` | Identify deviant case groups via decision tree |
| `unstable_params()` | Parameter instability analysis |
| `group_rules()` | Extract decision rules for deviant groups |
| `competes()` | Show competing splits at tree nodes |
| `assess_nca()` | Necessary Condition Analysis for PLS-SEM |
| `assess_nca_esse()` | NCA with Effect Size Sensitivity Extension |
| `assess_cta()` | Confirmatory Tetrad Analysis (CTA-PLS) |
| `assess_fimix()` | FIMIX-PLS latent class segmentation |
| `assess_fimix_compare()` | Compare FIMIX solutions across K values |
| `assess_pos()` | PLS-POS prediction-oriented segmentation |
| `assess_pos_compare()` | Compare PLS-POS solutions across K values |
| `pos_segments()` | Extract segment-specific re-estimated PLS models |
| `congruence_test()` | Bootstrapped congruence coefficient testing |

## Textbook demos

seminrExtras bundles the demo files for the *PLS-SEM using R* workbook (Hair
et al., 2026). After loading the package, list and run them with `demo()`:

```{r echo=TRUE, eval=FALSE}
demo("seminr-pls-cvpat", package = "seminrExtras")
```

Available demos include method walkthroughs (`seminr-pls-cvpat`,
`seminr-pls-pcm`, `seminr-pls-cipma`, `seminr-pls-coa`, `seminr-pls-nca`,
`seminr-pls-fimix`, `seminr-pls-cta`, `seminr-pls-pos`,
`seminr-pls-congruence`) and per-chapter primers
(`seminr-primer-v2-chap2` through `seminr-primer-v2-chap8`).

## Links

- **Source:** <https://github.com/sem-in-r/seminrExtras>
- **CRAN:** <https://cran.r-project.org/package=seminrExtras>
- **Issues:** <https://github.com/sem-in-r/seminrExtras/issues>

## References

Becker, J.-M., Richter, N. F., Ringle, C. M., & Sarstedt, M. (2026). Must-have, or maybe not? A sensitivity-based extension to necessary condition analysis. *Journal of Business Research, 206*, 115920. <https://doi.org/10.1016/j.jbusres.2025.115920>

Danks, N. P. (2021). The piggy in the middle: The role of mediators in PLS-SEM-based prediction. *ACM SIGMIS Database: The DATABASE for Advances in Information Systems, 52*(SI), 24–42. <https://doi.org/10.1145/3505639.3505644>

Danks, N. P., Ray, S., & Shmueli, G. (2024). The composite overfit analysis framework: Assessing the out-of-sample generalizability of construct-based models using predictive deviance, deviance trees, and unstable paths. *Management Science, 70*(1), 647–669. <https://doi.org/10.1287/mnsc.2023.4705>

Dul, J. (2016). Necessary condition analysis (NCA): Logic and methodology of "necessary but not sufficient" causality. *Organizational Research Methods, 19*(1), 10–52. <https://doi.org/10.1177/1094428115584005>

Franke, G. R., Sarstedt, M., & Danks, N. P. (2021). Assessing measure congruence in nomological networks. *Journal of Business Research, 130*, 318–334. <https://doi.org/10.1016/j.jbusres.2021.03.003>

Gudergan, S. P., Ringle, C. M., Wende, S., & Will, A. (2008). Confirmatory tetrad analysis in PLS path modeling. *Journal of Business Research, 61*(12), 1238–1249. <https://doi.org/10.1016/j.jbusres.2008.01.012>

Hair, J. F., Hult, G. T. M., Ringle, C. M., Sarstedt, M., Danks, N. P., & Ray, S. (2026). *Partial least squares structural equation modeling (PLS-SEM) using R: A workbook* (2nd ed.). Springer.

Hauff, S., Richter, N. F., Sarstedt, M., & Ringle, C. M. (2024). Importance and performance in PLS-SEM and NCA: Introducing the combined importance-performance map analysis (cIPMA). *Journal of Retailing and Consumer Services, 78*, 103723. <https://doi.org/10.1016/j.jretconser.2024.103723>

Liengaard, B. D., Sharma, P. N., Hult, G. T. M., Jensen, M. B., Sarstedt, M., Hair, J. F., & Ringle, C. M. (2021). Prediction: Coveted, yet forsaken? Introducing a cross-validated predictive ability test in partial least squares path modeling. *Decision Sciences, 52*(2), 362–392. <https://doi.org/10.1111/deci.12445>

Ringle, C. M., & Sarstedt, M. (2016). Gain more insight from your PLS-SEM results: The importance-performance map analysis. *Industrial Management & Data Systems, 116*(9), 1865–1886. <https://doi.org/10.1108/IMDS-10-2015-0449>

Sharma, P. N., Liengaard, B. D., Hair, J. F., Sarstedt, M., & Ringle, C. M. (2023). Predictive model assessment and selection in composite-based modeling using PLS-SEM: Extensions and guidelines for using CVPAT. *European Journal of Marketing, 57*(6), 1662–1677. <https://doi.org/10.1108/EJM-08-2020-0636>
